Long- and short-read metabarcoding technologies reveal similar spatio-temporal structures in fungal communities

### 2020-03-02 writing/transect_paper.Rmd
PERMANOVA and PCoA

An additional sample was collected at every sampling location (1-m spacing) in 2016 using LifeGuard^TM^ Soil Preservation Solution (MO BIO, Carlsbad, CA; USA) for preservation, without field lysis.
Sequencing results for these samples differed significantly (p < 0.001 from permutational ANOVA) from samples preserved using the Xpedition^TM^ lysis buffer (Supplementary Figures \@ref(fig:buffer-compare-long) and \@ref(fig:buffer-compare-short)); as such these samples were excluded from our spatial analyses.
However, reads from these sequences were included in the full bioinformatics workflow, including ASV calling, OTU clustering, and phylogenetic trees.

```{r}
buffer_asv_table <- rdd(proto_physeq) %>%
  phyloseq::subset_samples(tech != "Ion Torrent") %>%
  phyloseq::subset_samples(sample_type == "Sample") %>%
  phyloseq::prune_taxa(taxa = rdd(fungi_PHYLOTAX.Cons)$label) %>%
        phyloseq::prune_samples(samples = rowSums(phyloseq::otu_table(.)) > 100)

samples <- phyloseq::sample_data(buffer_asv_table) %>%
  as_tibble(rownames = "sample") %>%
  group_by(site, x, tech, amplicon) %>%
  filter(n() >= 3) %>%
  pull(sample) %>%
  unique()

buffer_asv_table <- phyloseq::prune_samples(buffer_asv_table, samples = samples)

buffer_asv_bc_dist <- buffer_asv_table %>%
  phyloseq::transform_sample_counts(function(x) x / sum(x)) %>%
  phyloseq::distance("bray")
vegan::adonis2(
  buffer_asv_bc_dist ~ paste(site, x) + paste(tech, amplicon) + buffer + year,
  data = as_tibble(phyloseq::sample_data(buffer_asv_table)),
  by = "margin",
  permutations = 9999
)
```

```{r}
buffer_asv_dbrda <- vegan::dbrda(
  buffer_asv_bc_dist ~ 1,
  data = as_tibble(phyloseq::sample_data(buffer_asv_table))
)
plot(buffer_asv_dbrda, type = "none")
points(buffer_asv_dbrda, select = with(as_tibble(phyloseq::sample_data(buffer_asv_table)),
                                     buffer == "Xpedition" & year == "2015" ),
       col = "green", pch = 1)
points(buffer_asv_dbrda, select = with(as_tibble(phyloseq::sample_data(buffer_asv_table)),
                                     buffer == "Xpedition" & year == "2016"),
       col = "blue", pch = 2)
points(buffer_asv_dbrda, select = with(as_tibble(phyloseq::sample_data(buffer_asv_table)),
                                     buffer == "LifeGuard" & year == "2016"),
       col = "red", pch = 3)
legend("bottomright", legend = c("Xpedition 2015", "Xpedition 2016", "LifeGuard 2016"), col = c("green", "blue", "red"), pch = 1:3)
legend("topright", legend = with(as_tibble(phyloseq::sample_data(buffer_asv_table)), paste(tech, amplicon)) %>% unique() %>% sort(), lty = 1:3)
vegan::ordiellipse(buffer_asv_dbrda, groups = with(as_tibble(phyloseq::sample_data(buffer_asv_table)), paste(tech, amplicon)),
                   lty = 1:3)
```

```{r}
tech_asv_table <- rdd(proto_physeq) %>%
  phyloseq::subset_samples(tech != "Ion Torrent") %>%
  phyloseq::subset_samples(buffer == "Xpedition") %>%
  phyloseq::subset_samples(sample_type == "Sample") %>%
  phyloseq::prune_taxa(taxa = rdd(fungi_PHYLOTAX.Cons)$label) %>%
  phyloseq::prune_samples(samples = rowSums(phyloseq::otu_table(.)) > 100)

tech_class_table <- phyloseq::`tax_table<-`(
  tech_asv_table,
  phyloseq::tax_table(
    rdd(fungi_PHYLOTAX.Cons) %>%
      select(label, kingdom:genus) %>%
      column_to_rownames("label") %>%
      as.matrix()
    )
  ) %>%
  phyloseq::tax_glom(taxrank = "class")

phyloseq::taxa_names(tech_class_table) <- phyloseq::tax_table(tech_class_table)[,"class"]

samples <- phyloseq::sample_data(tech_class_table) %>%
  as_tibble(rownames = "sample") %>%
  group_by(site, x, year) %>%
  filter(n() >= 3) %>%
  pull(sample) %>%
  unique()

tech_class_table <-
  phyloseq::prune_samples(tech_class_table, samples = samples) %>%
  phyloseq::transform_sample_counts(function(x) x / sum(x)) %>%
  phyloseq::prune_taxa(taxa = (phyloseq::otu_table(tech_class_table) %>% t() %>% as.data.frame() %>% do.call(what = pmax)) > 0.01)

tech_class_bc_dist <- tech_class_table %>%
  phyloseq::distance("bray")
vegan::adonis2(
  tech_class_bc_dist ~ paste(site, x, year) + tech + amplicon,
  data = as_tibble(phyloseq::sample_data(tech_class_table)),
  by = "margin",
  permutations = 9999
)
```

@@

(ref:ppcoa-short) Partial Principal Coordinates Analysis (PPCoA) of fungal community using different metabarcoding approaches.

(ref:ppcoa) Only within-sample variation is displayed; variation between samples (site, position, year) have been partialed out.

```{r ppcoa}
plot(tech_class_dbrda, type = "none")
text(
  x = tech_class_dbrda_scores$MDS1,
  y = tech_class_dbrda_scores$MDS2,
  labels = tech_class_dbrda_scores$abbrev
  )
points(tech_class_dbrda, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "PacBio" & amplicon == "Short" ),
       col = "green", pch = 1)

points(tech_class_dbrda, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "PacBio" & amplicon == "Long"),
       col = "blue", pch = 2)
points(tech_class_dbrda, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "Illumina" & amplicon == "Short"),
       col = "red", pch = 3)
legend("bottomleft", legend = c("Illumina Short", "PacBio Short", "PacBio Long"), col = c("red", "green", "blue"), pch = c(3, 1, 2))
```

@@

plot(tech_class_nmds, type = "n")
vegan::ordispider(
  tech_class_nmds,
  groups = with(
    as_tibble(phyloseq::sample_data(tech_class_table)),
    paste(buffer, year, site, x)
    ),
  col = "gray80"
  )
text(
  x = class_scores$MDS1,
  y = class_scores$MDS2,
  labels = class_scores$abbrev
  )
points(tech_class_nmds, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "PacBio" & amplicon == "Short" ),
       col = "green", pch = 1)
points(tech_class_nmds, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "PacBio" & amplicon == "Long"),
       col = "blue", pch = 2)
points(tech_class_nmds, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "Illumina" & amplicon == "Short"),
       col = "red", pch = 3)
legend("topright", legend = c("Illumina Short", "PacBio Short", "PacBio Long"), col = c("red", "green", "blue"), pch = c(3, 1, 2))
```

@@

>>>>

An additional sample was collected at every sampling location (1-m spacing) in 2016 using LifeGuard^TM^ Soil Preservation Solution (MO BIO, Carlsbad, CA; USA) for preservation, without field lysis.
Sequencing results for these samples differed significantly (p < `r rdd(buffer_adonis)["buffer","p.value"] %>% formatC(format = "f")` from permutational MANOVA, $R^2 = `r rdd(buffer_adonis)["buffer","R2"] %>% formatC(format = "f", digits = 2)`$ ) from samples preserved using the Xpedition^TM^ lysis buffer (Supplementary Figures \@ref(fig:buffer-pcoa), \@ref(fig:buffer-compare-long), and \@ref(fig:buffer-compare-short)); as such these samples were excluded from our spatial analyses.
However, reads from these samples were included in the full bioinformatics workflow, including ASV calling, OTU clustering, and phylogenetic trees.

```{r}

@@

(ref:ppcoa-short) Partial Principal Coordinates Analysis (PPCoA) of fungal community using different metabarcoding approaches

(ref:ppcoa) Spatio-temporal variation (i.e., variation between different soil samples) has been partialled out.
Only samples with at least 100 reads for all three approaches are included.
ASVs are clustered by taxonomic class.
Abbreviations represent the approximate direction of increase for the classes with the highest scores: `r glue::glue_data(tech_class_dbrda_scores, "**{abbrev}**: {class}") %>% glue::glue_collapse(sep = "; ", last = "; and ")`.

```{r ppcoa, fig.cap = "(ref:ppcoa-short). (ref:ppcoa)", fig.scap = "(ref:ppcoa-short)"}
plot(tech_class_dbrda, type = "none")
points(tech_class_dbrda, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "PacBio" & amplicon == "Short" ),
       col = "green", pch = 1)

points(tech_class_dbrda, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "PacBio" & amplicon == "Long"),
       col = "blue", pch = 2)
points(tech_class_dbrda, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "Illumina" & amplicon == "Short"),
       col = "red", pch = 3)
text(
  x = tech_class_dbrda_scores$MDS1,
  y = tech_class_dbrda_scores$MDS2,
  labels = tech_class_dbrda_scores$abbrev
  )
legend("bottomleft", legend = c("Illumina Short", "PacBio Short", "PacBio Long"), col = c("red", "green", "blue"), pch = c(3, 1, 2))
```

@@

tech_class_nmds_scores <-
  vegan::scores(tech_class_nmds, display = "species") %>%
  as_tibble(rownames = "class") %>%
  mutate(L = sqrt(NMDS1^2 + NMDS2^2)) %>%
  filter(!is.nan(L)) %>%
  filter(L >= max(L) / 2) %>%
  mutate(abbrev = substr(class, start = 1, stop = 2))

plot(tech_class_nmds, type = "n")
vegan::ordispider(
  tech_class_nmds,
  groups = with(
    as_tibble(phyloseq::sample_data(tech_class_table)),
    paste(buffer, year, site, x)
    ),
  col = "gray80"
  )
text(
  x = tech_class_nmds_scores$NMDS1,
  y = tech_class_nmds_scores$NMDS2,
  labels = tech_class_nmds_scores$abbrev
  )
points(tech_class_nmds, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "PacBio" & amplicon == "Short" ),
       col = "green", pch = 1)
points(tech_class_nmds, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "PacBio" & amplicon == "Long"),
       col = "blue", pch = 2)
points(tech_class_nmds, select = with(as_tibble(phyloseq::sample_data(tech_class_table)),
                                     tech == "Illumina" & amplicon == "Short"),
       col = "red", pch = 3)
legend("topright", legend = c("Illumina Short", "PacBio Short", "PacBio Long"), col = c("red", "green", "blue"), pch = c(3, 1, 2))
```

@@

(ref:buffer-pcoa-short) Unconstrained Principal Coordinate Analysis ordination of samples preserved with different buffers

(ref:buffer-pcoa) For each sequencing strategy, only those sampling locations with more than 100 reads each for all three year/preservation buffer combinations were included.

```{r buffer-pcoa, fig.cap="(ref:buffer-pcoa-s). (ref:buffer-pcoa)"}
plot(rdd(buffer_asv_pcoa), type = "none")
points(
  rdd(buffer_asv_pcoa),
  select = with(
    as_tibble(phyloseq::sample_data(rdd(buffer_asv_table))),
    buffer == "Xpedition" & year == "2015" 
  ),
  col = "green",
  pch = 1
)
points(
  rdd(buffer_asv_pcoa),
  select = with(as_tibble(phyloseq::sample_data(rdd(buffer_asv_table))),
                buffer == "Xpedition" & year == "2016"
  ),
  col = "blue",
  pch = 2
)
points(
  rdd(buffer_asv_pcoa),
  select = with(as_tibble(phyloseq::sample_data(rdd(buffer_asv_table))),
                buffer == "LifeGuard" & year == "2016"
  ),
  col = "red",
  pch = 3
)
legend(
  "bottomright",
  legend = c("Xpedition 2015", "Xpedition 2016", "LifeGuard 2016"),
  col = c("green", "blue", "red"),
  pch = 1:3
)
# legend(
#   "topright",
#   legend = with(as_tibble(phyloseq::sample_data(rdd(buffer_asv_table))), paste(tech, amplicon)) %>%
#     unique() %>%
#     sort(),
#   lty = 1:3
# )
# vegan::ordiellipse(
#   rdd(buffer_asv_pcoa),
#   groups = with(as_tibble(phyloseq::sample_data(rdd(buffer_asv_table))), paste(tech, amplicon)),
#   lty = 1:3
# )
```
